Distortion-corrected kernel density estimate on Riemannian manifolds

### 2022-07-23 paper/paper.tex
Sec 2 done

For multivariate data, variable kernel density estimator with adaptive bandwidth matrix at each sample point can be used to capture the local structure of the data. \textcite{Pelletier2005-vu} generalize the multivariate kernel density esitmator to Riemnnian manifolds. However, it is not applicable when the underlying manifold is not known or when the true manifold is mapped in a much higher dimensional space. Therefore, we propose a distortion-corrected kernel density estimator as in Equation \eqref{eq:denestimator0} based on the low-dimensional embedding, \(\pmb{y}_i \in \mathbb{R}^d, i = 1, 2, \dots, N\), of the high-dimensional data with \(N\) data points,
\begin{equation}
\label{eq:denestimator0}
\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}} K\bigg( \frac{ \sqrt{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)} }{r} \bigg),
\end{equation}
where \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\) is the pointwise adaptive Riemannian matrix of dimension \(d \times d\).
The proposed estimator could be used as an estimate of the density of the true manifold, with the local distortion at each point corrected by volume density functions. These densities could further be used for unsupervised tasks including anomaly detection when outliers are defined as the points with the lowest densities.

To start with, we introduce the multivariate kernel density estimation method with variable bandwidth matrix and Pelletier's estimator for Riemannian manifolds. Then we propose our distortion-corrected estimator, followed by details about how to estimate the Riemannian metric using the Learn Metric algorithm in \textcite{Perrault-Joncas2013-pq}. Finally, we show how to use these densities to carrt out anomaly detection tasks for high dimensional data that lies on a Riemannian manifold densities.

@@

Multivariate kernel density estimation has gained lots of attention in exploratory data analysis. It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data. \textcite{Scott2015-vl} covers a wide range of techniques for estimating the kernel densities of multivariate data.
In general, given a multivariate random sample \(\pmb{X}_1, \dots, \pmb{X}_N \in \mathcal{R}^d\)
drawn from a density \(f\), the variable kernel density estimate \autocite{Breiman1977-qc,Jones1990-oe,Terrell1992-ut} of \(f\) at point \(\pmb{x}=(x_1, x_2, \dots, x_d)^\prime \in \mathcal{R}^d\) is given by
\begin{equation}
\label{eq:vkde}
\hat{f}(\pmb{x};\pmb{H}_i)=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i).
\end{equation}
where \(K(\pmb{x})\) is the kernel function, \(\pmb{H}_i\) is called the bandwidth matrix and it is a symmetric positive definite matrix of dimension \(d\times d\), and \(K_{\pmb{H}_i}(\pmb{x}) = |\pmb{H}_i|^{-1/2} K(\pmb{H}_i^{-1/2}\pmb{x})\).
When using the standard gaussian kernel, we have \(K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})\) and hence
\begin{equation}
\label{eq:kernel}
K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}_i^{-1}(\pmb{x}-\pmb{X}_i)\right],
\end{equation}
with \(|\pmb{H}_i|\) being the determinant of matrix \(\pmb{H}_i\).
Intuitively, KDE smooths out each data point \(\pmb{X}_i\) into a smooth bump with the shape determined by the kernel function. Then an average over all data points is taken as the estimated density \(\hat{f}\). KDE is a local smooth estimator because the nearby points contributes more to the density value compared to points far away from the estimated data point \(\pmb{x}\).

@@

Consider a compact Riemannian manifold \((M, g)\) of dimension \(d\) without boundary and a probability distribution with density \(f\) on the manifold. Assume \((M, d_g)\) is a complete metric space, where \(d_g\) is the geodesic distance induced by the Riemannian metric \(g\) and the strictly positive injectivity radius of the manifold \(inj_g(M)\) \autocite{Chavel2006-mp}. Denote \(\pmb{X}_i, i = 1, 2, \dots, N\), where \(\pmb{X}_i \in \mathcal{R}^d\), as i.i.d. random objects on \(M\) with density \(f\). To estimate the density of the manifold, \textcite{Pelletier2005-vu} proposed a kernel density estimator for \(f\) at point \(\pmb{p}\in M\) as
\begin{equation}
\label{eq:denriem}
\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{\pmb{X}_i}(\pmb{p})} K(\frac{d_g(\pmb{p}, \pmb{X}_i)}{r}),
\end{equation}
where \(r\) is the global bandwidth satisfying \(r \leq r_0\) for fixed \(r_0\) such that \(0\leq r_0\leq inj_g(M)\), \(K: \mathcal{R}_{+} \rightarrow \mathcal{R}\) is a non-negative map, and \(\theta_{\pmb{X}_i}(\pmb{p})\) is the volume density function on the manifold for point \(\pmb{X}_i\). The estimator in \eqref{eq:denriem} is also proved to be consistent with the kernel density estimators in the Euclidean case and converges at the same rate as the Euclidean estimator \autocite{Henry2009-ll}.

@@

For the exponential map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form \(\exp_p^*\textit{Vol}\) on \(T_pM\), \(\theta_p\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_pM\), which gives
\[
d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du,
\]
in normal coordinates \autocite{Le_Brigant2019-lj}. The volume density function is motivated by how a measure \(du\) changes under the change of coordinates. The volume form is variant to the exponential map except that the manifold is flat. Therefore, a multiplicative corrected term called the volume density function has to be added in the estimator so that it will integral to 1. It is closely related to the integration on manifolds and plays an important part in studying the density of manifolds.
\textcite{Pelletier2005-vu} mentions that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) expressed in the coordinates at the logarithm map of q, \(\exp_q^{-1}(p)\).
Note that this is true only when using the geodesic normal coordinates. If the mapping function is not exponential/logarithm map, the volume density function is calculated otherwise. Therefore, the Riemannian manifold density estimator in \textcite{Pelletier2005-vu} only works when the true manifold \(M\) is known.
Further, the fixed bandwidth assumes that the smoothing is the same across the entire manifold and there is no local distortion at each sample point, which needs to be corrected using an adaptive bandwidth matrix when distortion exists.

@@

Motivated by the estimator in \eqref{eq:denriem}, we restate below our proposed distortion-corrected kernel density estimator (DC-KDE) of \(f\), same as \eqref{eq:denestimator0}, based on the manifold embedding \(\pmb{y}_i, i = 1, 2, \dots, N\),
\begin{equation}
\label{eq:denestimator}
\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}} K\bigg( \frac{ \sqrt{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)} }{r} \bigg),
\end{equation}
where \(r\) is a global bandwidth scalar, \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\), is the pointwise variable Riemannian matrix and it is positive definite with dimension \(d\times d\), \(\frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}}\) is the estimated volume density function when using the coordinates from any manifold learning mapping, \(\sqrt{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}\) is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric \(\pmb{H}(\pmb{y}_i)\), and \(C\) is a correction term for the volume between the spaces so that the integral is 1. We will now state below how to calculate each term in the estimator. Then the only problem left is the estimation of the Riemannian metric at each embedding point, to be discussed in \autoref{riemmetric}.

@@

Now we consider the case with the manifold learning embedding. Consider the chart \((U, \psi)\) induced by an orthonormal basis of the tangent space of the embedding \(T_{\pmb{y}_i}E\) and a normal neighborhood of \(\pmb{y}_i\). Given a query point \(\pmb{p} \in B(\pmb{y}_i,r)\), the volume of the parallelepiped is spanned by the same basis.\\
Then the volume density function is given by
\begin{equation}
\label{eq:volumedensityest}
\theta_{\pmb{y}_i}(\pmb{p}) = \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}},
\end{equation}
where \(\pmb{H}\) is the estimated Riemannian metric for the unknown manifold.
Again, the volume density function is not defined for all \(\pmb{p}\) but for points within the geodesic ball satisfying \(\pmb{p} \in B(\pmb{y}_i,r)\), which is related to the support of \(K\) being \([0,1]\) in condition (v) of \eqref{eq:kernelcondition}.

As for the geodesic distance \(d_g\) in \eqref{eq:denriem}, we propose a distortion-fixed geodesic distance with the use of estimated Riemannian metric as
\begin{equation}
\label{eq:dgestimate}
\hat{d}_g(\pmb{p}, \pmb{y}_i) = \sqrt{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)},
\end{equation}
where \(\pmb{H}(\pmb{y}_i)\) is the estimated Riemannian metric at embedding point \(\pmb{y}_i\).

For two points \(p,q \in M\), the geodesic distance between \(p\) and \(q\) is defined as the length of the shortest curve from \(p\) to \(q\) along the manifold.
The geodesic distance is intrinsic, which means it is a coordinate-independent quantity. The Riemannian metric functions as the correction term for different coordinates of manifold learning embedding.
By using the first-order distance estimator, \(\hat{d}_g(\pmb{p}, \pmb{y}_i)\) serves as a good approximation of the true geodesic distance \(d_g\) with the distortion corrected at each embedding point \(\pmb{y}_i\). Compared with the graph distance, the shortest path between point in the neighborhood graph \(G\), our proposed \(\hat{d}_g\) is better because we use the same graph \(G\) for all manifold learning embeddings but the distances between different nodes in \(G\) might not be the same due to the distortion.

As pointed out by \textcite{Perrault-Joncas2013-pq}, the naive distance \(\|p-q\|\) underestimates \(d_g\), and the graph shortest distance \(d_G\) overestimates \(d_g\) by the triangle inequality. By using only the first-order correction in \(\hat{d}_g\), our estimator is also affected by the estimated Riemannian metric. In order to correct for the estimation error in \(\pmb{H}(\pmb{y}_i)\), we further divide the density estimator in \eqref{eq:denestimator} by a constant term \(C\) so that the integral of the kernel is equal to 1. The constant term in \(1\)-d is given by the \emph{area under the curve (AUC)}, and can be named as the \emph{volume between the surface (VBS)} in higher dimension. We use the volume estimator in Section 6.4 of \textcite{Perrault-Joncas2013-pq} to sum over all embedding points \(\pmb{y}_i\) and get
\begin{equation}
\label{eq:vbs}
C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i),
\end{equation}
where the distortion is corrected for the tangent plane of the embedding points.

@@

The Riemannian metric \(g\) is a symmetric and positive definite tensor field which defines an inner product \(\langle,\rangle_g\) on the tangent space \(T_pM\) for every point \(p \in M\). If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of the manifold. For each point \(p\in M\) in any embedding coordinate system, the Riemannian metric \(g\) is a \(d\times d\) symmetric positive definite matrix, where \(d\) is the dimension of the manifold.

The inner product between two vectors \(u,v \in T_pM\), given by \(\langle u,v \rangle_g=g_{ij}u^iv^j\){[}\^{}1{]}, can be used to define some geometric quantities, such as the vector norm \(|u|=\sqrt{\langle u,u \rangle_g}\) and the angle between two vectors \(\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}\) in the tangent space.
{[}\^{}1{]}: Here the Einstein notation is used where superscripts denote summation over \(i\) and \(j\).
The line element and volume element of the full manifold or a subset of the manifold can also be computed from \(g\). The arc length of a curve \(c\) along the manifold is defined as
\[
l(c)=\int_a^b \sqrt{g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt}} dt,
\]
where \((x^1,\dots,x^d)\) are the coordinates of chart \((U,x)\) and \(c(t)\) is a function that maps \([a,b]\) to \(M\). While the volume of \(V\subset M\) is computed by
\[
Vol(V)=\int_V \sqrt{|\det g|} dx^1\dots dx^d.
\]
Both the concepts of distance and volume are relevant to kernel density estimation as we have shown in Equation \eqref{eq:volumedensityratio}, \eqref{eq:volumedensityest}, \eqref{eq:dgestimate}, and \eqref{eq:vbs}.

@@

As pointed out by \textcite{Perrault-Joncas2013-pq}, if the embedding dimension is larger than the manifold intrinsic dimension \(d\), the rank of the embedding metric \(\pmb{H}(p)\) is equal to \(d\); otherwise, the Riemannian metric will be returned.
To ensure that the estimated Riemannian metric is returned as a \(d\times d\) matrix, for unknown manifold dimension, we apply the intrinsic dimension (id) selection algorithm called \emph{two-nearest neighbor estimator (TWO-NN estimator)} \autocite{Facco2017-rl,Denti2021-jl} to derive id estimator from a Poisson point process. \textcite{Facco2017-rl} point out that of the distances between a point and its second and first neareast neighbors follow a Pareto distribution with scale parameter \(1\) and shape parameter \(d\), the intrinsic dimension of the manifold \autocite{Facco2017-rl}. By fitting a Pareto distribution to proper transformed data, the intrinsic dimension could be estimated. An \emph{R} library named \emph{intRinsic} has implemented this algorithm and we are going to apply it to when \(d\) is unknown.

\begin{algorithm}[!htb]
  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
  \label{alg:learnmetric}
  \DontPrintSemicolon
  \SetAlgoLined
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
  \Input{ high-dimensional data $\pmb{x}_i \in \mathbb{R}^s$ for all $i=1,\ldots,N$ }
  \Output{ low-dimensional data $\pmb{y}_i \in \mathbb{R}^d$ and its Riemannian metric $\pmb{H}(\pmb{y}_i)$ for all $i=1,\ldots,N$ }
  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
  \OptParameter{ manifold learning parameters $MANILEARN$}
  \BlankLine
  \begin{algorithmic}[1]

@@

Now we are able to generate kernel density estimates of the underlying manifold using the estimator in Equation \eqref{eq:denestimator}. The densities could be further applied to unsupervised learning tasks such as clustering and anomaly detection. As one way to utilize the kernel density estimates, we propose an algorithm for anomaly detection in \ref{alg:kdeanomaly}.
There are mainly four steps involved in the algorithm. The first step is to apply the Learn metric algorithm described in \ref{alg:learnmetric} to the input high-dimensional data \(x_i,i=1,\ldots,N\) where \(x_i\in \mathcal{R^s}\) to get the low-dimensional embedding \(y_i\), where \(y_i\in \mathcal{R^d}\) and the Riemannian metric \(\pmb{H}_i\) with dimension \(d\times d\) for each observation. Two parameters \(\sqrt{\varepsilon} = 0.4\) and \(c=0.25\) are given for Gaussian kernels. Then we could use the pointwise Riemannian metric to calculate the kernel density estimate with our proposed estimator in Equation \eqref{eq:denestimator}. The top outliers of size \(n\_outliers\) are obtained by ordering the embedding points \(y_i\) according to their density estimates \(f_N(\pmb{y}_i)\).

@@

{\centering \includegraphics[width=0.95\linewidth]{figures/kde_h-vkde}

>>>>

For multivariate data, variable kernel density estimator with adaptive bandwidth matrix at each sample point can be used to capture the local structure of the data. \textcite{Pelletier2005-vu} generalize the multivariate kernel density esitmator to Riemnnian manifolds. However, it is not applicable when the underlying manifold is unknown or when the manifold is mapped in a much higher dimensional space. Therefore, we propose a distortion-corrected kernel density estimator as in Equation \eqref{eq:denestimator0} based on the low-dimensional manifold learning embedding, \(\pmb{y}_i \in \mathbb{R}^d, i = 1, 2, \dots, N\), of the high-dimensional data,
\begin{equation}
\label{eq:denestimator0}
\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
\end{equation}
where \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\) is the pointwise adaptive Riemannian matrix of dimension \(d \times d\).
The proposed estimator could be used as an estimate of the true manifold density, with the local distortion corrected by volume density function at each point. These densities could further be used for unsupervised tasks including anomaly detection when outliers are defined as the points with the lowest densities.

To start with, we introduce the multivariate kernel density estimation method with variable bandwidth matrix and Pelletier's kernel density estimator for Riemannian manifolds. Then we propose our distortion-corrected estimator, followed by details about how to estimate the Riemannian metric using the Learn Metric algorithm from \textcite{Perrault-Joncas2013-pq}. Finally, we show how to use these densities to carrt out anomaly detection tasks for high dimensional data that lies on Riemannian manifolds.

@@

Multivariate kernel density estimation has gained lots of attention in exploratory data analysis. It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data. \textcite{Scott2015-vl} covers a wide range of techniques for estimating the kernel densities of multivariate data.
In general, given multivariate random samples \(\pmb{X}_1, \dots, \pmb{X}_N \in \mathbb{R}^d\)
drawn from a density \(f\), the variable kernel density estimate \autocite{Breiman1977-qc,Jones1990-oe,Terrell1992-ut} of \(f\) at point \(\pmb{x}=(x^1, x^2, \dots, x^d)^\prime \in \mathbb{R}^d\) is given by
\begin{equation}
\label{eq:vkde}
\hat{f}(\pmb{x};\pmb{H}_i)=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i).
\end{equation}
where \(K(\pmb{x})\) is the kernel function, \(\pmb{H}_i\) is called the bandwidth matrix and it is a symmetric positive definite matrix of dimension \(d\times d\), and \(K_{\pmb{H}_i}(\pmb{x}) = |\pmb{H}_i|^{-1/2} K(\pmb{H}_i^{-1/2}\pmb{x})\).
When using the standard gaussian kernel, we have \(K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})\) and hence
\begin{equation}
\label{eq:kernel}
K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}_i^{-1}(\pmb{x}-\pmb{X}_i)\right],
\end{equation}
with \(|\pmb{H}_i|\) being the determinant of matrix \(\pmb{H}_i\).
Intuitively, KDE smooths out each data point \(\pmb{X}_i\) into a smooth bump with the shape determined by the kernel function. Then an average over all data points is taken as the estimated density \(\hat{f}\). KDE is a local smooth estimator because the nearby points contributes more to the density value compared to points far away from the estimated data point \(\pmb{x}\).

@@

Consider a compact Riemannian manifold \((M, g)\) of dimension \(d\) without boundary and a probability distribution with density \(f\) on the manifold. Assume \((M, d_g)\) as a complete metric space, where \(d_g\) is the geodesic distance induced by the Riemannian metric \(g\) and the strictly positive injectivity radius of the manifold \(inj_g(M)\) \autocite{Chavel2006-mp}. Denote \(\pmb{X}_i, i = 1, 2, \dots, N\), where \(\pmb{X}_i \in \mathbb{R}^d\), as i.i.d. random objects on \(M\) with density \(f\). To estimate the density of the manifold, \textcite{Pelletier2005-vu} propose a kernel density estimator for \(f\) at point \(\pmb{p}\in M\) as
\begin{equation}
\label{eq:denriem}
\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{\pmb{X}_i}(\pmb{p})} K(\frac{d_g(\pmb{p}, \pmb{X}_i)}{r}),
\end{equation}
where \(r\) is the global bandwidth satisfying \(r \leq r_0\) for fixed \(r_0\) such that \(0\leq r_0\leq inj_g(M)\), \(K: \mathbb{R}_{+} \rightarrow \mathbb{R}\) is a non-negative map, and \(\theta_{\pmb{X}_i}(\pmb{p})\) is the volume density function on the manifold for point \(\pmb{X}_i\). The estimator in \eqref{eq:denriem} is also proved to be consistent with the kernel density estimators in the Euclidean case and converges at the same rate as the Euclidean estimator \autocite{Henry2009-ll}.

@@

For the exponential map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form \(\exp_p^*\textit{Vol}\) on \(T_pM\), \(\theta_p\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_pM\), which gives
\[
d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du,
\]
in normal coordinates \autocite{Le_Brigant2019-lj}. The volume density function is motivated by how a measure \(du\) changes under the change of coordinates. The volume form is variant to the exponential map except that the manifold is flat. Therefore, a multiplicative corrected term called the volume density function has to be added in the estimator so that it will integral to 1. It is closely related to the integration on manifolds and plays an important part in studying the density of manifolds.

\textcite{Pelletier2005-vu} mention that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) expressed in the coordinates at the logarithm map of q, \(\exp_q^{-1}(p)\).
Note that this is true only when using the geodesic normal coordinates. If the mapping function is not exponential/logarithm map, the volume density function is calculated otherwise. Therefore, the Riemannian manifold density estimator in \textcite{Pelletier2005-vu} only works when the true manifold \(M\) is known.
Further, the fixed bandwidth assumes that the smoothing is the same across the entire manifold and there is no local distortion at each sample point, which needs to be corrected using an adaptive bandwidth matrix when distortion exists.

@@

Motivated by the estimator in \eqref{eq:denriem}, we restate below our proposed distortion-corrected kernel density estimator (DC-KDE) of \(f\), same as \eqref{eq:denestimator0}, based on the manifold learning embedding \(\pmb{y}_i, i = 1, 2, \dots, N\),
\begin{equation}
\label{eq:denestimator}
\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
\end{equation}
where \(r\) is a global bandwidth scalar, \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\), is the pointwise variable Riemannian matrix and it is positive definite with dimension \(d\times d\), \(\bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2}\) is the estimated volume density function when using the coordinates from any manifold learning mapping, \(\pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)\) is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric \(\pmb{H}(\pmb{y}_i)\), and \(C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i)\) is a correction term for the volume between the spaces so that the integral is 1. We will now state below how to calculate each term in the estimator. And the only problem left is the estimation of the Riemannian metric at each embedding point, to be discussed in \autoref{riemmetric}.

@@

Now we consider the case with the manifold learning embedding. Consider the chart \((U, \psi)\) induced by an orthonormal basis of the tangent space of the embedding \(T_{\pmb{y}_i}E\) and a normal neighborhood of \(\pmb{y}_i\). Given a query point \(\pmb{p} \in B(\pmb{y}_i,r)\), the volume of the parallelepiped is spanned by the same basis.\\
Then the volume density function is given by
\begin{equation}
\label{eq:volumedensityest}
\theta_{\pmb{y}_i}(\pmb{p}) = \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2},
\end{equation}
where \(\pmb{H}\) is the estimated Riemannian metric for the unknown manifold.
Again, the volume density function is not defined for all \(\pmb{p}\) but for points within the geodesic ball satisfying \(\pmb{p} \in B(\pmb{y}_i,r)\), which is related to the support of \(K\) being \([0,1]\) in Condition (v) of \eqref{eq:kernelcondition}.

As for the geodesic distance \(d_g\) in \eqref{eq:denriem}, we propose a distortion-fixed geodesic distance with the use of estimated Riemannian metric as
\begin{equation}
\label{eq:dgestimate}
\hat{d}_g(\pmb{p}, \pmb{y}_i) = \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i),
\end{equation}
where \(\pmb{H}(\pmb{y}_i)\) is the estimated Riemannian metric at embedding point \(\pmb{y}_i\).

For two points \(\pmb{p},\pmb{q} \in M\), the geodesic distance between \(\pmb{p}\) and \(\pmb{q}\) is defined as the length of the shortest curve from \(\pmb{p}\) to \(\pmb{q}\) along the manifold.
The geodesic distance is intrinsic, which means it is a coordinate-independent quantity. The Riemannian metric functions as the correction term for different coordinates of manifold learning embedding.
By using the first-order distance estimator, \(\hat{d}_g(\pmb{p}, \pmb{y}_i)\) serves as a good approximation of the true geodesic distance \(d_g\) with the distortion corrected at each embedding point \(\pmb{y}_i\). Compared with the graph distance, the shortest path between point in the neighborhood graph \(G\), our proposed \(\hat{d}_g\) is better because we use the same graph \(G\) for all manifold learning embeddings but the distances between different nodes in \(G\) might not be the same due to the distortion.

As pointed out by \textcite{Perrault-Joncas2013-pq}, the naive distance \(\|\pmb{p}-\pmb{q}\|\) underestimates \(d_g\), and the graph shortest distance \(d_G\) overestimates \(d_g\) by the triangle inequality. By using only the first-order correction in \(\hat{d}_g\), our estimator is also affected by the estimated Riemannian metric. In order to correct for the estimation error in \(\pmb{H}(\pmb{y}_i)\), we further divide the density estimator in \eqref{eq:denestimator} by a constant term \(C\) so that the integral of the kernel is equal to 1. The constant term in \(1\)-d is given by the \emph{area under the curve (AUC)}, and can be named as the \emph{volume between the surface (VBS)} in higher dimension. We use the volume estimator in Section 6.4 of \textcite{Perrault-Joncas2013-pq} to sum over all embedding points \(\pmb{y}_i\) and get
\begin{equation}
\label{eq:vbs}
C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i),
\end{equation}
where the distortion is corrected for the tangent plane of the embedding points.

@@

The Riemannian metric \(g\) is a symmetric and positive definite tensor field which defines an inner product \(\langle,\rangle_g\) on the tangent space \(T_{\pmb{p}} M\) for every point \(\pmb{p} \in M\). If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of the manifold. For each point \(\pmb{p} \in M\) in any embedding coordinate system, the Riemannian metric \(g\) is a \(d\times d\) symmetric positive definite matrix, where \(d\) is the dimension of the manifold.

The inner product between two vectors \(u,v \in T_{\pmb{p}}M\), given by \(\langle u,v \rangle_g=g_{ij}u^iv^j\){[}\^{}1{]}, can be used to define some geometric quantities, such as the vector norm \(|u|=\sqrt{\langle u,u \rangle_g}\) and the angle between two vectors \(\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}\) in the tangent space.
{[}\^{}1{]}: Here the Einstein notation is used where superscripts denote summation over \(i\) and \(j\).

The line element and volume element of the full manifold or a subset of the manifold can also be computed from \(g\). The arc length of a curve \(c\) along the manifold is defined as
\[
l(c)=\int_a^b \sqrt{g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt}} dt,
\]
where \((x^1,\dots,x^d)\) are the coordinates of chart \((U,x)\) and \(c(t)\) is a function that maps \([a,b]\) to \(M\). While the volume of \(V\subset M\) is computed by
\[
Vol(V)=\int_V \sqrt{|\det g|} dx^1\dots dx^d.
\]
Both the concepts of distance and volume are relevant to kernel density estimation as we have shown in Equation \eqref{eq:volumedensityratio}, \eqref{eq:volumedensityest}, \eqref{eq:dgestimate}, and \eqref{eq:vbs}.

@@

As pointed out by \textcite{Perrault-Joncas2013-pq}, if the embedding dimension is larger than the manifold intrinsic dimension \(d\), the rank of the embedding metric \(\pmb{H}(p)\) is equal to \(d\); otherwise, the Riemannian metric will be returned.
To ensure that the estimated Riemannian metric is returned as a \(d\times d\) matrix, for unknown manifold dimension, we apply the intrinsic dimension (id) selection algorithm called \emph{two-nearest neighbor estimator (TWO-NN estimator)} \autocite{Facco2017-rl,Denti2021-jl} to derive id estimator from a Poisson point process. \textcite{Facco2017-rl} point out that of the distances between a point and its second and first neareast neighbors follow a Pareto distribution with scale parameter \(1\) and shape parameter \(d\), the intrinsic dimension of the manifold \autocite{Facco2017-rl}. By fitting a Pareto distribution to proper transformed data, the intrinsic dimension could be estimated. An \emph{R} library named \emph{intRinsic} \autocite{Denti2021-qc} has implemented this algorithm and we are going to apply it to when \(d\) is unknown.

\begin{algorithm}[!htb]
  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
  \label{alg:learnmetric}
  \DontPrintSemicolon
  \SetAlgoLined
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
  \Input{ high-dimensional data $\pmb{x}_i \in \mathbb{R}^s$ for all $i=1,\ldots,N$ }
  \Output{ low-dimensional data $\pmb{y}_i \in \mathbb{R}^d$ and its Riemannian metric $\pmb{H}(\pmb{y}_i)$ for all $i=1,\ldots,N$ }
  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
  \OptParameter{ manifold learning parameters EMBED }
  \BlankLine
  \begin{algorithmic}[1]

@@

Now we are able to generate kernel density estimates of the underlying manifold using the estimator in Equation \eqref{eq:denestimator}. The densities could be further applied to unsupervised learning tasks such as clustering and anomaly detection. As one way to utilize the kernel density estimates, we propose an algorithm for anomaly detection in \ref{alg:kdeanomaly}.
There are mainly four steps involved in the algorithm. The first step is to apply the Learn metric algorithm described in \ref{alg:learnmetric} to the input high-dimensional data \(x_i,i=1,\ldots,N\) where \(x_i\in \mathcal{R^s}\) to get the low-dimensional embedding \(y_i\), where \(y_i\in \mathcal{R^d}\) and the Riemannian metric \(\pmb{H}_i\) with dimension \(d\times d\) for each observation. Two parameters \(\sqrt{\varepsilon} = 0.4\) and \(c=0.25\) are given for Gaussian kernels. Then we could use the pointwise Riemannian metric to calculate the kernel density estimate with our proposed estimator in Equation \eqref{eq:denestimator}. The top outliers of size \(n\_outliers\) are obtained by ordering the embedding points \(y_i\) according to their density estimates \(f(\pmb{y}_i)\).

@@

{\centering \includegraphics[width=0.95\linewidth]{figures/DC-KDE}
